Comparison of Terrain Following and Cut Cell Grids with a Non-Hydrostatic Model

### 2015-04-13 msc-publication.tex
revert the mountain profile equation to include h^\star, fix typos

\title{Comparison of Terrain Following and Cut Cell Grids with a Non-Hydrostatic Model}
\authors{James Shaw\correspondingauthor{Dept., Institution, Address, City, State/Country.}}
\affiliation{}
\email{j.shaw@pgr.reading.ac.uk}

%but are known to suffer from horizontal pressure gradient errors near steep terrain.
\abstract{Terrain following coordinates are widely used in operational models.
More recently, the cut cell method has been proposed as an alternative that can more accurately represent orography.  In order to compare the accuracy of the different grids, it is typically necessary to use different models.  Instead, we use a single model to enable a like-for-like comparison of terrain following and cut cell grids.
A series of standard two-dimensional tests using idealised terrain are performed: tracer advection in a prescribed horizontal velocity field, a test of a stably stratified atmosphere at rest, and orographically induced gravity waves.  In addition, we formulate a new tracer advection test having a velocity field that is tangential to the basic terrain following coordinate surfaces.
The results of the tracer advection tests demonstrate that tracer accuracy depends upon alignment of the flow with the grid.  In the gravity waves test, results on all grids are in good agreement with existing results from the literature.  However, we find that the Lorenz computational mode is manifested as a vertical zig--zag in potential temperature only on the cut cell grid.
Although errors do vary between grids, overall, a high degree of accuracy is obtained for all tests on all grids.}

@@

\section{Introduction}
Representing orography accurately in numerical weather prediction systems is necessary to model downslope winds and local precipitation patterns.  There are two main approaches to represent orography on a grid: terrain following layers and cut cells.  Both methods align cells in vertical columns.  Existing studies of cut cells have compared results with tests on terrain following grids from previous studies that used different models.  On the contrary, this study uses a single model to enable a like-for-like comparison between terrain following and cut cell grids.

@@

Cut cells is an alternative method in which cells that lie entirely below the terrain are removed, and those that intersect the surface are modified in shape so that they more closely fit the terrain.  The resulting grid is orthogonal everywhere except where cells have been cut.  Hence, errors are still introduced when calculating the horizontal pressure gradient between cut and uncut cells.

@@

Several studies have found that cut cells produce more accurate results when compared to TF coordinates.  Spurious winds seen in TF coordinates are not present and errors do not increase with steeper terrain \citep{good2014}.  \TODO{anything else?}

This study uses the nonhydrostatic model from \citet{weller-shahrokhi2014} to enable a like-for-like comparison between terrain following and cut cell grids for idealised, two-dimensional test cases from the literature.  Section 2 presents the formulation of the terrain following and cut cell grids used in the experiments that follow.  Section 3 analyses the results from two tracer advection tests, a test of a stratified atmosphere at rest, and orographically induced gravity waves.  Concluding remarks are made in section 4.

\begin{figure*}
	\centering
	\includegraphics{../msc-publication-fig-meshes/fig-meshes.pdf}
	%
	\caption{Examples of (a) BTF, (b) SLEVE, and (c) a cut cell grid, all constructed with OpenFOAM.  The two dimensional grids are \SI{20}{\kilo\meter} wide and \SI{2}{\kilo\meter} high.  The cut cell grid was created by intersecting the terrain surface with a regular grid as described in section~\ref{sec:grid}.}
	\label{fig:grid}
\end{figure*}

@@

Here we describe the formulation of the terrain following grids and the method of cut cell grid construction.  The techniques presented here are used to define the grids for the experiments in the subsequent section.

\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h \label{eqn:btf}
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the height of the coordinate surface at level \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.

@@

The smooth level vertical (SLEVE) coordinate proposed by \citet{schaer2002} achieves a more regular TF grid in the middle and top of the domain than the BTF coordinate.  The terrain height is split into large-scale and small-scale components, \(h_1\) and \(h_2\), such that \(h = h_1 + h_2\), with each component having a different exponential decay. The transformation is defined as 
\begin{align}
	z &= z^\star + h_1 b_1 + h_2 b_2
\intertext{where the vertical decay functions are given by}
	b_i &= \frac{\sinh \left( \left( H / s_i \right)^n - \left( z^\star / s_i \right)^n \right)}{\sinh \left( H / s_i \right)^n}
\end{align}
with \(s_1\) and \(s_2\) are the scale heights of large-scale and small-scale terrain respectively.  The exponent \(n\) was introduced by \citet{leuenberger2010} in order to increase cell thickness in the layers nearest the ground, allowing longer timesteps and permitting more accurate calculation of parameterised low-level heat and momentum fluxes.  \citet{leuenberger2010} found the exponent has an optimal value of \(n = 1.35\).  Choosing \(n = 1\) gives the decay functions used by \citet{schaer2002}.

@@

\section{Results}
A series of standard, two-dimensional tests are performed over idealised orography.  For each test, results on the BTF, SLEVE and cut cell grid are compared.  The first test from \citet{schaer2002} advects a tracer in a horizontal velocity field.  Second, a new tracer advection test is formulated employing a terrain following velocity field to challenge the advection scheme on more orthogonal grids.  The third test solves the Euler equations for a stably stratified atmosphere at rest, following \citet{klemp2011}.  Finally, as specified by \citet{schaer2002}, a test of orographically-induced gravity waves is performed.

@@

Following \citet{schaer2002}, a tracer is transported above wave-shaped terrain by solving the advection equation for a prescribed horizontal wind.  This challenges the accuracy of the advection scheme in the presence of grid distortions.

The domain is \SI{300}{\kilo\meter} wide and \SI{25}{\kilo\meter} high which is discretized onto a grid with \(\Delta x = \SI{1}{\kilo\meter}\) and \(\Delta z^\star = \SI{500}{\meter}\).  The terrain is wave-shaped, specified by the surface height, \(h\), such that
\begin{align}
	h(x) &= \left\{ \begin{array}{l l}
	h_0 \cos^2 ( \alpha x ) \cos^2 ( \beta x ) & \quad \text{if $| x | < a$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.
\end{align}
where $a = \SI{25}{\kilo\meter}$ is the mountain half-width, $h_0 = \SI{3}{\kilo\meter}$ is the maximum mountain height, $\lambda = \SI{8}{\kilo\meter}$ is the wavelength, \(\alpha = \pi / \lambda\) and \(\beta = \pi / 2a\).  On the SLEVE grid, the large-scale component $h_1$ is given by \(h_1(x) = h^\star(x) / 2\)
and $s_1 = \SI{15}{\kilo\meter}$ is the large scale height, and $s_2 = \SI{2.5}{\kilo\meter}$ is the small scale height.  The optimisation of SLEVE by \citet{leuenberger2010} is not used, so the exponent $n = 1$.

@@

A tracer $\varphi$ is positioned upstream above the height of the terrain.  It has the shape
\begin{align}
	\varphi(x, z) &= \varphi_0 \left\{ \begin{array}{l l}
		\cos^2 \left( \frac{\pi r}{2} \right) & \quad \text{if $r \leq 1$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.
%
\intertext{having radius $r$ given by}
%
	r &= \sqrt{
		\left( \frac{x - x_0}{A_x} \right)^2 + 
		\left( \frac{z - z_0}{A_z} \right)^2
	}
\end{align}
where $A_x = \SI{25}{\kilo\meter}$, $A_z = \SI{3}{\kilo\meter}$ are the horizontal and vertical half-widths respectively, and $\varphi_0 = 1$ is the maximum magnitude of the anomaly.  At $t = \SI{0}{\second}$, the anomaly is centred at $(x_0, z_0) = (\SI{-50}{\kilo\meter}, \SI{9}{\kilo\meter})$ so that the anomaly is upwind of the mountain and well above the maximum terrain height of \SI{3}{\kilo\meter}.  Analytic solutions can be found by adjusting the anomaly centre such that $x_0 = ut$.

@@

The test was executed on the BTF, SLEVE and cut cell grids, and on a regular grid with flat terrain.  Tracer contours at \(t = \SI{0}{\second}, \SI{5000}{\second}\) and \(\SI{10000}{\second}\) are shown in Figure~\ref{fig:advection-tracer}.  The result from \citet{schaer2002} using a fourth-order centred scheme and sigma coordinates (\ref{fig:advection-tracer}a) is compared with the upwind-biased cubic scheme on (\ref{fig:advection-tracer}b) the BTF grid, and (\ref{fig:advection-tracer}c) cut cell grid.

@@

Minimum and maximum tracer values and \(\ell^2\) error norms on the BTF, SLEVE, cut cell and regular grids are summarised in table~\ref{tab:advection}.  The results of the cubic upwind-biased scheme on TF and regular grids are comparable with those for the fourth-order centred scheme from \citet{schaer2002}.  Error is largest on the BTF grid with \(\ell^2 = \num{0.00758}\) but significant reduced on the SLEVE grid with \(\ell^2 = \num{0.00110}\).  The error is approximately halved by changing from the SLEVE grid to the cut cell grid.

@@

\begin{figure}
	\centering
	\includegraphics{../msc-publication-fig-advection-tracer/fig-advection-tracer.pdf}
%
	\caption{Horizontally advected tracer contours at \(t = \SI{0}{\second}\), \SI{5000}{\second} and \SI{10000}{\second} using (a) the fourth-order centred difference scheme with sigma coordinates from \citet{schaer2002}, and the upwind-biased cubic scheme on (b) BTF grid, (c) cut cell grid.  Contour intervals are every 0.1.  The terrain profile is also shown immediately above the ground.}
	\label{fig:advection-tracer}
\end{figure}

@@

We conclude from this test that accuracy depends upon alignment of the flow with the grid, and accuracy is not significantly reduced by grid distortions.  Accuracy on the BTF grid is comparable with accuracy on the SLEVE grid in the horizontal advection test.

\subsection{Resting atmosphere}
An idealised terrain profile is defined along with a stratified atmosphere at rest in hydrostatic balance.  The analytic solution is time-invariant, but numerical errors in calculating the horizontal pressure gradient can give rise to spurious velocities which become more severe over steeper terrain \citep{klemp2011}.

@@

The wave-shaped mountain profile has a surface height, $h$, given by
\begin{align}
	h(x) = h_0 \exp \left( - \left( \frac{x}{a} \right)^2 \right) \cos^2 \left( \frac{\pi x}{\lambda} \right) \label{eqn:resting:mountain}
\end{align}
where $a = \SI{5}{\kilo\meter}$ is the mountain half-width, $h_0 = \SI{1}{\kilo\meter}$ is the maximum mountain height and $\lambda = \SI{4}{\kilo\meter}$ is the wavelength.  For the optimised SLEVE grid, the large-scale component $h_1$ is specified as
\begin{align}
h_1(x) = \frac{1}{2} h_0 \exp \left( - \left( \frac{x}{a} \right)^2 \right)
\end{align}
and, following \cite{leuenberger2010}, $s_1 = \SI{4}{\kilo\meter}$ is the large scale height, $s_2 = \SI{1}{\kilo\meter}$ is the small scale height, and the optimal exponent value of $n = 1.35$ is used.

@@

Sponge layers are added to the upper \SI{10}{\kilo\meter} and leftmost \SI{10}{\kilo\meter} at the inlet boundary to damp the reflection of waves.
The term $\mu \rho \bf{u}$ is subtracted from the momentum equation \TODO{include the equation set in this paper, or cite weller-shahrokhi2014 here?} where the damping function $\mu$ is adapted from \citet{melvin2010} such that
\begin{align}
	\mu(x, z) &= \mu_\mathrm{upper} + \mu_\mathrm{inlet} \\
	\mu_\mathrm{upper}(z) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{z - z_B}{H - z_B} \right) & \text{if } z \geq z_B \\
		0 & \text{otherwise} \\
	\end{cases} \\
	\mu_\mathrm{inlet}(x) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{x_I - x}{x_I - x_0} \right) & \text{if } x < x_I \\
		0 & \text{otherwise}
	\end{cases}
\end{align}
where $\overline{\mu} = 1.2$ is the damping coefficient, $z_B = \SI{20}{\kilo\meter}$ is the bottom of the sponge layer, $H = \SI{30}{\kilo\meter}$ is the top of the domain, $x_0 = \SI{-150}{\kilo\meter}$ is the leftmost limit of the domain and $x_I = \SI{-140}{\kilo\meter}$ is the rightmost extent of the inlet sponge layer.  The sponge layer is only active on faces whose normal is vertical so that it damps vertical momentum only.

Note that, while the domain itself is \SI{30}{\kilo\meter} in height, for the purposes of generating of BTF and SLEVE grids, the domain height is set to \SI{20}{\kilo\meter} because the sponge layer occupies the uppermost \SI{10}{\kilo\meter}.

@@

\begin{figure}
	\centering
	\includegraphics{../msc-publication-fig-gravityWaves/fig-gravityWaves.pdf}
%
	\caption{Vertical cross section of vertical velocity contours in the gravity waves test after 5 hours on (a) the BTF grid compared with (b) the mass-conserving semi-implicit semi-Lagrangian solution from \citet{melvin2010}.  Contours are every \SI{5e-2}{\meter\per\second} with solid lines denoting ascent and dashed lines descent.}
	\label{fig:gw-w}
\end{figure}

>>>>

\title{Comparison of Terrain Following and Cut Cell Grids with a Non-Hydrostatic Model}
\authors{James Shaw\correspondingauthor{Department of Meteorology, University of Reading, Earley Gate, PO Box 243, Reading, RG6 6BB, UK.}}
\affiliation{}
\email{j.shaw@pgr.reading.ac.uk}

%but are known to suffer from horizontal pressure gradient errors near steep terrain.
\abstract{Terrain following coordinates are widely used in operational models but, more recently, the cut cell method has been proposed as an alternative that can more accurately represent the orography and associated atmospheric dynamics.  Because the type of grid is usually chosen during model implementation, it is typically necessary to use different models in order to compare the accuracy of different grids.  On the contrary, we use a single model to enable a like-for-like comparison of terrain following and cut cell grids.
A series of standard two-dimensional tests using idealised terrain are performed: tracer advection in a prescribed horizontal velocity field, a test of a stably stratified atmosphere at rest, and orographically induced gravity waves.  In addition, we formulate a new tracer advection test having a velocity field that is tangential to the basic terrain following coordinate surfaces.
The results of the advection tests demonstrate that tracer accuracy depends upon alignment of the flow with the grid.  In the gravity waves test, results on all grids are in good agreement with existing results from the literature.  However, we find that the Lorenz computational mode is manifested as a vertical zig--zag in potential temperature only on the cut cell grid.
Although errors do vary between grids, overall, a high degree of accuracy is obtained for all tests on all grids.}

@@

\section{Introduction}
Representing orography accurately in numerical weather prediction systems is necessary to model downslope winds and local precipitation patterns.  There are two main approaches to represent orography on a grid: terrain following layers and cut cells.  Both methods align cells in vertical columns.  Because most models are designed for a particular type of grid, existing studies of cut cells have compared results with terrain following grids using different models.  Instead, this study uses a single model to enable a like-for-like comparison between terrain following and cut cell grids.

@@

Cut cells is an alternative method in which cells that lie entirely below the terrain are removed, and those that intersect the surface are modified in shape so that they more closely fit the terrain.  The resulting grid is orthogonal everywhere except near cells that have been cut.  Hence, errors are still introduced when calculating the horizontal pressure gradient between cut and uncut cells.

@@

Several studies have found that cut cells produce more accurate results when compared to TF coordinates.  Spurious winds seen in TF coordinates are not present and errors do not increase with steeper terrain \citep{good2014}.  \TODO{any other studies?}

This study uses the nonhydrostatic model from \citet{weller-shahrokhi2014} to enable a like-for-like comparison between terrain following and cut cell grids for idealised, two-dimensional test cases from the literature.  Section 2 presents the formulation of the terrain following and cut cell grids used in the experiments that follow.  Section 3 analyses the results from two tracer advection tests, a test of a stably stratified atmosphere at rest, and orographically induced gravity waves.  Concluding remarks are made in section 4.

\begin{figure*}
	\centering
	\includegraphics{../msc-publication-fig-meshes/fig-meshes.pdf}
	%
	\caption{Examples of (a) BTF, (b) SLEVE, and (c) a cut cell grid, all constructed with OpenFOAM.  The two dimensional grids are \SI{20}{\kilo\meter} wide and \SI{2}{\kilo\meter} high.  SLEVE parameters are specified in the resting atmosphere test in section~\ref{sec:results}\ref{sec:resting}.  The cut cell grid was created by intersecting the terrain surface with a regular grid as described in section~\ref{sec:grid}.}
	\label{fig:grid}
\end{figure*}

@@

Here we describe the formulation of the terrain following grids and the method of cut cell grid construction.  The techniques presented are used to define the grids for the experiments in the subsequent section.

\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h \label{eqn:btf}
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the height of the coordinate surface at level \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.  An example is shown in figure~\ref{fig:grid}a.

@@

The smooth level vertical (SLEVE) coordinate proposed by \citet{schaer2002} achieves a more regular TF grid in the middle and top of the domain than the BTF coordinate.  The terrain height is split into large-scale and small-scale components, \(h_1\) and \(h_2\), such that \(h = h_1 + h_2\), with each component having a different exponential decay. The transformation is defined as 
\begin{align}
	z &= z^\star + h_1 b_1 + h_2 b_2
\intertext{where the vertical decay functions are given by}
	b_i &= \frac{\sinh \left( \left( H / s_i \right)^n - \left( z^\star / s_i \right)^n \right)}{\sinh \left( H / s_i \right)^n}
\end{align}
with \(s_1\) and \(s_2\) are the scale heights of large-scale and small-scale terrain respectively.  The exponent \(n\) was introduced by \citet{leuenberger2010} in order to increase cell thickness in the layers nearest the ground, allowing longer timesteps and permitting more accurate calculation of parameterised low-level heat and momentum fluxes.  \citet{leuenberger2010} found the exponent has an optimal value of \(n = 1.35\).  Choosing \(n = 1\) gives the decay functions used by \citet{schaer2002}.  An example of the SLEVE grid can be seen in figure~\ref{fig:grid}b.

@@

\section{Results}
\label{sec:results}

A series of standard, two-dimensional tests are performed over idealised orography.  For each test, results on the BTF, SLEVE and cut cell grid are compared.  The first test from \citet{schaer2002} advects a tracer in a horizontal velocity field.  Second, a new tracer advection test is formulated employing a terrain following velocity field to challenge the advection scheme on more orthogonal grids.  The third test solves the Euler equations for a stably stratified atmosphere at rest, following \citet{klemp2011}.  Finally, as specified by \citet{schaer2002}, a test of orographically-induced gravity waves is performed.

@@

Following \citet{schaer2002}, a tracer is transported above wave-shaped terrain by solving the advection equation for a prescribed horizontal wind.  This test challenges the accuracy of the advection scheme in the presence of grid distortions.

The domain is \SI{300}{\kilo\meter} wide and \SI{25}{\kilo\meter} high, discretized onto a grid with \(\Delta x = \SI{1}{\kilo\meter}\) and \(\Delta z^\star = \SI{500}{\meter}\).  The terrain is wave-shaped, specified by the surface height, \(h\), such that
\begin{subequations}
\begin{align}
   h(x) &= h^\star \cos^2 ( \alpha x )
%
\intertext{where}
%
   h^\star(x) &= \left\{ \begin{array}{l l}
       h_0 \cos^2 ( \beta x ) & \quad \text{if $| x | < a$} \\
	0 & \quad \text{otherwise}
    \end{array} \right.
\end{align}
\end{subequations}
where $a = \SI{25}{\kilo\meter}$ is the mountain half-width, $h_0 = \SI{3}{\kilo\meter}$ is the maximum mountain height, $\lambda = \SI{8}{\kilo\meter}$ is the wavelength, \(\alpha = \pi / \lambda\) and \(\beta = \pi / 2a\).  On the SLEVE grid, the large-scale component $h_1$ is given by \(h_1(x) = h^\star(x) / 2\)
and $s_1 = \SI{15}{\kilo\meter}$ is the large scale height, and $s_2 = \SI{2.5}{\kilo\meter}$ is the small scale height.  The optimisation of SLEVE by \citet{leuenberger2010} is not used, so the exponent $n = 1$.

@@

A tracer, $\varphi$, is positioned upstream above the height of the terrain.  It has the shape
\begin{align}
	\varphi(x, z) &= \varphi_0 \left\{ \begin{array}{l l}
		\cos^2 \left( \frac{\pi r}{2} \right) & \quad \text{if $r \leq 1$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.
%
\intertext{having radius, $r$, given by}
%
	r &= \sqrt{
		\left( \frac{x - x_0}{A_x} \right)^2 + 
		\left( \frac{z - z_0}{A_z} \right)^2
	}
\end{align}
where $A_x = \SI{25}{\kilo\meter}$, $A_z = \SI{3}{\kilo\meter}$ are the horizontal and vertical half-widths respectively, and $\varphi_0 = 1$ is the maximum magnitude of the anomaly.  At $t = \SI{0}{\second}$, the anomaly is centred at $(x_0, z_0) = (\SI{-50}{\kilo\meter}, \SI{9}{\kilo\meter})$ so that the anomaly is upwind of the mountain and well above the maximum terrain height of \SI{3}{\kilo\meter}.  Analytic solutions can be found by setting the anomaly centre such that $x_0 = ut$.

@@

The test was executed on the BTF, SLEVE and cut cell grids, and on a regular grid with flat terrain.  Tracer contours at \(t = \SI{0}{\second}, \SI{5000}{\second}\) and \(\SI{10000}{\second}\) are shown in Figure~\ref{fig:advection-tracer}.  The result from \citet{schaer2002} using a fourth-order centred scheme and sigma coordinates (\ref{fig:advection-tracer}a) is compared with the upwind-biased cubic scheme on the BTF grid (\ref{fig:advection-tracer}b), and cut cell grid (\ref{fig:advection-tracer}c).

@@

Minimum and maximum tracer values and \(\ell^2\) error norms on the BTF, SLEVE, cut cell and regular grids are summarised in table~\ref{tab:advection}.  The results of the cubic upwind-biased scheme on TF and regular grids are comparable with those for the fourth-order centred scheme from \citet{schaer2002}.  Error is largest on the BTF grid with \(\ell^2 = \num{0.00767}\) but significant reduced on the SLEVE grid with \(\ell^2 = \num{0.00108}\).  The error is approximately halved by changing from the SLEVE grid to the cut cell grid.

@@

\begin{figure}
	\centering
	\includegraphics{../msc-publication-fig-advection-tracer/fig-advection-tracer.pdf}
%
	\caption{Horizontally advected tracer contours at \(t = \SI{0}{\second}\), \SI{5000}{\second} and \SI{10000}{\second} using (a) the fourth-order centred difference scheme with sigma coordinates from \citet{schaer2002}, and the upwind-biased cubic scheme on (b) BTF grid, (c) cut cell grid.  Contour intervals are every 0.1.  The terrain profile is also shown immediately above the $x$ axis.}
	\label{fig:advection-tracer}
\end{figure}

@@

We conclude from this test that accuracy depends upon alignment of the flow with the grid, and accuracy is not significantly reduced by grid distortions.  Error on the BTF grid is comparable with the error on the SLEVE grid in the horizontal advection test.

\subsection{Resting atmosphere}
\label{sec:resting}

An idealised terrain profile is defined along with a stably stratified atmosphere at rest in hydrostatic balance.  The analytic solution is time-invariant, but numerical errors in calculating the horizontal pressure gradient can give rise to spurious velocities which become more severe over steeper terrain \citep{klemp2011}.

@@

The wave-shaped mountain profile has a surface height, $h$, given by
\begin{align}
	h(x) = h_0 \exp \left( - \left( \frac{x}{a} \right)^2 \right) \cos^2 \left( \alpha x \right) \label{eqn:resting:mountain}
\end{align}
where $a = \SI{5}{\kilo\meter}$ is the mountain half-width, $h_0 = \SI{1}{\kilo\meter}$ is the maximum mountain height and $\lambda = \SI{4}{\kilo\meter}$ is the wavelength.  For the optimised SLEVE grid, the large-scale component $h_1$ is specified as
\begin{align}
h_1(x) = \frac{1}{2} h_0 \exp \left( - \left( \frac{x}{a} \right)^2 \right)
\end{align}
and, following \cite{leuenberger2010}, $s_1 = \SI{4}{\kilo\meter}$ is the large scale height, $s_2 = \SI{1}{\kilo\meter}$ is the small scale height, and the optimal exponent value of $n = 1.35$ is used.

@@

Sponge layers are added to the upper \SI{10}{\kilo\meter} and leftmost \SI{10}{\kilo\meter} at the inlet boundary to damp the reflection of waves.
The term $\mu \rho \bf{u}$ is subtracted from the momentum equation \TODO{include the equation set in this paper, or cite weller-shahrokhi2014 here?} where the damping function $\mu$ is adapted from \citet{melvin2010} such that
\begin{align}
	\mu(x, z) &= \mu_\mathrm{upper} + \mu_\mathrm{inlet} \\
	\mu_\mathrm{upper}(z) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{z - z_B}{H - z_B} \right) & \text{if } z \geq z_B \\
		0 & \text{otherwise} \\
	\end{cases} \\
	\mu_\mathrm{inlet}(x) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{x_I - x}{x_I - x_0} \right) & \text{if } x < x_I \\
		0 & \text{otherwise}
	\end{cases}
\end{align}
where $\overline{\mu} = 1.2$ is the damping coefficient, $z_B = \SI{20}{\kilo\meter}$ is the bottom of the sponge layer, $H = \SI{30}{\kilo\meter}$ is the top of the domain, $x_0 = \SI{-150}{\kilo\meter}$ is the leftmost limit of the domain and $x_I = \SI{-140}{\kilo\meter}$ is the rightmost extent of the inlet sponge layer.  The sponge layer is only active on faces whose normal is vertical so that it damps vertical momentum only. \TODO{is this true of the inlet sponge as well?}

Note that, while the domain itself is \SI{30}{\kilo\meter} in height, for the purposes of generating BTF and SLEVE grids, the domain height is set to \SI{20}{\kilo\meter} because the sponge layer occupies the uppermost \SI{10}{\kilo\meter}.

@@

\begin{figure}
	\centering
	\includegraphics{../msc-publication-fig-gravityWaves/fig-gravityWaves.pdf}
%
	\caption{Vertical cross section of vertical velocity contours in the gravity waves test after 5 hours with mountain height $h_0 = \SI{250}{\meter}$ on (a) the BTF grid compared with (b) the mass-conserving semi-implicit semi-Lagrangian solution from \citet{melvin2010}.  Contours are every \SI{5e-2}{\meter\per\second} with solid lines denoting ascent and dashed lines descent.}
	\label{fig:gw-w}
\end{figure}
